Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-06-11 paper/paper.tex
new logo

% official NIPS keywords
\textbf{\\keywords}: mind-wandering, artificial intelligence, Markov decision process

@@

\section{Introduction}
%
In the absence of external stimulation, the human brain is not at rest.
In the beginning of the 21st century,
brain imaging may have been the first technique
to allow for the discovery of a unique brain network that probably maintains
baseline mental activities
\citep{raichle2001pnas, randy2008, bzdok2015resting}.
The ``default mode network'' (DMN) continues
to metabolize large quantities of
oxygen and glucose energy to maintain
neuronal computation
during free-ranging thought
\citep{kenet2003spontaneously, fiser2004small}.
The baseline energy demand is only weakly modulated
at the onset of defined psychological tasks \citep{raichle_baseline}.
At the opposite, during sleep, the
decoupling of brain structures discarded the idea of the DMN being
only a passive network resonance and rather supported an
important role in sustaining conscious awareness \citep{horovitz2009decoupling}.

This dark matter of brain physiology \citep{raichle2006brain} begs the question
of the biological purpose underlying DMN activity.
What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
We propose that this set of some of the most advanced regions in
the association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control of human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex'' \citep{carhart2010default}.

@@

Since theories based on the FEP \citep{friston2010free,fristonAIorRL}
conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
given the environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a light-deprived closed space can be used
  as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would readily
identify the visual-input-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.

@@

\section{Conclusion}
Which brain function could be important enough
for the existence and survival of the human species
to justify constantly high energy costs?
MDPs motivate an attractive
formal account how the human association cortex
might implement multi-sensory representation and control of the environment to
optimize the organism's interaction with the world.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the view of a Markovian sequential decision process,
human behavior unfolds by integrating action outcomes from stored past events
and extrapolation to
upcoming events for guiding action choice in the present context.
MDPs also provide a mathematical formalism how
opportunity structure in the environment can be recursively exploited
when confronted with challenging decisions.
%
This functional interpretation may be more compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world representation
by injecting stochasticity in the recall of past
actions and the estimation of action outcomes may explain why
highly accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.

The proposed artificiale intelligence perspective
on DMN biology is hence practically computable and
yields falsifiable neurobiological hypotheses.
%
Neuroscience experiments could be designed that operationalize
the set of action, value, and state variables that govern
the behavior of intelligent RL systems.
%
At the least, we propose an alternative vocabulary to
describe and interpret experimental findings in neuroscience studies
on the DMN situated at top of the brain network hierarchy.
%
Ultimately,
DMN activity may instantiate a holistic integration
ranging from real experience over purposeful dreams to anticipated futures
for continued refinement of the organism's fate.

@@

\appendix

\begin{remark}
  Given a system $(\mathcal S, \mathcal A, r, p, \mu)$ satisfying all the axioms for an MDP
  except the Markov property, we can always transform it into an MDP by considering a compounded
  version of the states
  $$S_t \leftarrow (S_0,A_0,R_0,S_1,A_1,S_2\ldots,S_t,A_{t-1},R_{t-1},S_{t})$$ made of the system's
  history up to an including time $t$.
\end{remark}

\section{More on RL in the brain}
\subsection{Does hippocampal replay equal inverse reinforcement-learning?}
%\begin{mdframed}
  Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
  an MDP (called a \textit{teacher's demonstration}), can we figure out what is the
  (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
  \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
  (e.g., linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
  For instance, given traces of motor actions from an adult to grab a cup on a table,
  can an observing child figure out what ``langragian'' functional is being minimized by the former?
  How? Can they reproduce this optimal behavior?
%\end{mdframed}
Such questions are of course pertinent to our decision-making theory for the DMN.
In the general artificial case, the problem has been extensively studied and partially solved by
\citep{abbeel2004}. They are also been rigorously studied in general optimal control literature under the name
``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...

IRL is is suited for problems in which it's hard to define what the reward function could be
(e.g.,  car-driving, drone maneuvering, etc.) ...

\section{Free-energy principles!}
The so-called free-energy principle in its present form (including notions like ``generative density'', ``recognition density'', etc.) can be traced back to works of Dayan \& Hinton \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial directional deep belief-net (DBN).
In this subsection, we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in building their Helmholtz machine. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For instance, the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial) wake-sleep algorithm to train the underlying Helmholtz machine, etc.

% \begin{mdframed}
%   Though our presentation of free-energy principles is thermodynamic in flavor, a completely equivalent
%   information-theoretic formulation is of course possible, simply by exploiting the information-theoretic interpretation of entropy.
% \end{mdframed}

\subsection{Helmholtz free-energy and the generative model}
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{11cm}}
         \hline
         \textbf{symbol}    & \textbf{description}  \\ \hline
         $\langle X\rangle_p$ & Expectation (a.k.a average, a.k.a mean value) of the
         random quantity $X$ w.r.t to the probability density $p$, formally defined by $\langle E\rangle_p := \sum_{z}p(z)X(z)$.\\ \hline
         $\mathcal H(p)$ & Information-theoretic entropy of a probability density $p$, formally defined by $\mathcal H(p) := -\sum_{z}p(z)\log(p(z))$,
          with the usual convention $0 \log(0) := 0$.\\ \hline
         $D_{KL}(q||p)$ & The Kullback-Leibler divergence between the probability densities $q$ and $p$ respectively, formally defined by $D_{KL}(q||p) := \sum_{z}q(z)\log(q(z)/p(z))$.\\ \hline
             $\x$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's internal state $b$ and sensory inputs $s$, i.e., $\x = (s, b).$ \\ \hline
             $\z$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
             $p_G(.|\x)$ & Generative density for ...\\ \hline
         $p_R(.|\x)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
         $F_G(\x)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\x$. This measures the surprise incured upon observing $\x$ generated by the model $G$.\\ \hline
         $F^R_G(\x)$ & Variational Helmholtz free-energy from model $G$
          to $R$.  Note that $F^G_G = F_G$.\\ \hline
  \end{tabular}
  \caption{Table of notations.}
\end{table}
Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
\begin{eqnarray}
  \begin{split}
    \text{generative surprise } &= -\log(p_G(\x)) = -\log(p_G(\x)) \times 1 = -\log(P_G(\x))\sum_{\z}p_G(\z |\x)\\
    &= -\sum_{\z}p_G(\z, \x)\log(p_G(\x))
    =-\sum_{\z}p_G(\z |\x)\log(p_G(\z, \x)/p_G(\z|\x))\\
    &= \sum_{\z}p_G(\z |\x)\log(p_G(\z|\x))-\sum_{\z}p_G(\z |\x)\log  (p_G(\z, \x))\\
    &= -\langle \log  (p_G(., \x)) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x))\\
    &= \langle E_G(., \x) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x))
  \end{split}
  \label{eq:helm}
\end{eqnarray}
where $E_G(\z, \x)$ is the energy at \textit{macrostate} $\z$ of a fictive thermodynamic system defined by setting
\begin{equation}
  E_G(\z, \x) := -\log(p_G(\z, \x)),
  \label{eq:gibbs}
\end{equation}
%% In the above, the variable $\z$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
The last quantity in \eqref{eq:helm} is nothing but \textit{Helmholtz free-energy} (at unit temperature!), defined formally by
\begin{equation}
  F_G(\x) := \langle E_G(., \x) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x)).
\end{equation}
%% , modelled as the physical system for which $p_G(\z|\x)$ dictates the occupation probabilities of macrostates $\z$
Thus,
\begin{fact}
  Generative surprise and generative Helmholtz free-energy are different views on exactly the same object.
\end{fact}

The goal of the brain is then to optimize over the generative model $G$: to iteratively or analytically modify the generative density $p_G(.|\x)$, so as to minimize surprise. It turns out that a direct attempt to attack this optimization problem by gradient descent on the free-energy $F_G(\x)$ is furtile: the parameter update steps are not ``very clean'', and require rather cumbersome and heavy computations. A workaround is then to introduce a second density $p_R(.|\x)$ called a \textit{recognition} density to work in tandem with the generative density $p_G(.|\x)$, as a trick for doing approximate inference. The former dreams / fantacizes whilst the latter tries to generate sensations which match these dreams! This primal-dual idea, first proposed in Hinton et al. 1995, is at the heart of the general free-energy principle that we will introduce shortly.

\subsection{Variational Helmholtz free-energy and the bottom-up recognition sub-model}
In this subsection, we will present an insightful upper bound for the generative surprise (i.e., generate Helmholtz free-energy), called the \textit{variational} (Helmholtz) free-energy. As an avant-gout of what is to come shorty, let's just note that the well-known \textit{free-energy principle} is simply a workaround whereby the minimization surprise (intractable) is replaced with the  minimization a carefully chosen upper bound thereof.

Invoking \eqref{eq:gibbs} and applying Bayes rule, we get the Gibbs
distribution
\begin{equation}
  p_G(\z|\x) = \frac{p_G(\z|\x)}{p_G(\x)} = \frac{\exp(-E_G(\z, \x))}{Z_G(\x)} = \frac{\exp(-E_G(\z, \x))}{Z_G(\x)},
\end{equation}
where $Z_G(\x) := \log(p_G(\x)) = \sum_{\z'}\exp(-E_G(\z',\x))$, the normalizing \textit{partition function} for the model \ref{eq:gibbs}.
Whence, for any macrostate $\z$, we have $p_G(\x) = Z_G(\x) = \exp(-E_G(\z, \x)) / p_G(\z|\x)$, and so it holds that
\begin{equation}
F_G(\x)\overset{\eqref{eq:helm}}{=} -\log(p_G(\x)) = -\log(Z_G(\x)) = E_G(\z, \x) + \log(p_G(\z|\x)).
\end{equation}
Now, in the above equation, the LHS only depends on the generative model $G$ and the data point $\x$:
it doesn't depend on the hidden variable $\z$, etc. So, taking expectations w.r.t an arbitrary
density\footnote{conditioning in $P_R(.|\x)$ is because this density is selected from a world in which the sensory inputs and internal brain
  state vector $\x$ is assumed already observed.} $p_R(.|\x)$ yields
\begin{equation}
  \begin{split}
    F_G(\x) &= -\log(Z_G(\x)) = \langle E_G(., \x)\rangle_{P_R(.|\x)} + \sum_{\z}p_R(\z|\x)\log(p_G(\z|\x))\\
    &= \langle E_G(., \x)\rangle_{P_R(.|\x)} - \mathcal H(p_R(.|\x)) - \sum_{\z}p_R(\z|\x)\log(p_R(\z|\x)/p_G(\z|\x))\\
    &= F^R_G(\x) - D_{KL}(P_R(.|\x) || P_G(.|\x)),
  \end{split}
  \label{eq:fe}
\end{equation}
where $F^R_G(\x)$ is the \textit{variational} Helmholtz free-energy from $R$ to $G$ defined by
\begin{equation}
  F^R_G(\x) := \langle E_G(., \x)\rangle_{P_R(.|\x)} - \mathcal H(p_R(.|\x))
\end{equation}
and $D_{KL}(P_R(.|\x) || P_G(.|\x))$ is the Kullback-Leibler divergence between the $p_R(.|\x)$ and the generative density $p_G(.|\x)$. Note that $F^G_G = F_G$.

\subsection{A general free-energy principle}
We can resume the situation as follows\footnote{Where we have used the fact that KL divergence is always nonnegative.}:
\begin{mdframed}
\begin{equation}
  \begin{split}
    \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
    &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{D_{KL}(P_R(.|\x) || P_G(.|\x))}_{\text{complexity}} \\
    &\le F^R_G(\x),
    \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z
    \end{split}
\end{equation}
\end{mdframed}
%% In fact, we have the following theorem
%% \begin{theorem}
%%   It holds that
%%   $$
%%   \text{Minimal generative surprise = minimal (Helmholtz) variational free-energy},$$
%%   i.e
%%   \begin{equation}
%%     \min_{G}F_G(\x) = \min_{G,R}F^R_G(\x)
%%     \end{equation}
%% \end{theorem}

\subsection{Helmholtz machines and the wake-sleep algorithm}
% \begin{mdframed}
  \textbf{Assumption:} In both generative and recognition components of the network, there is conditional independence of  neurons in the same layer, given the data (i.e., input from lower more primitive layers). Precisely
$$
p_G(\z^{(l)}|\x) = \Pi_{k=1}^{h_l}p_G(\z_k^{(l)} | \x),\hspace{1cm}
p_R(\z^{(l)}|\x) = \Pi_{k=1}^{h_l}p_R(\z_k^{(l)} |\x)
$$
% \end{mdframed}

\subsection{Friston's active-inference and agency}
\label{sec:friston}
This is nothing but an application of the Dayan's wake-sleep algorithm for training a Helmholtz machine model of the brain...

The following critics can be made:
\begin{itemize}
  \item
    As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
  \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
  \item Also, note that the wake-sleep algorithm doesn't do backprop! This is due to
    technical difficulty in getting derivatives of loss function w.r.t
    recognition weights $\W^R$).
  \item This difficulty was removed in the 2010s by \citep{kingma2013auto}, an other groups, via a ``reparametrization trick''.
    \end{itemize}

\subsection{Minimizing free-energy via backprop: variational auto-encoders}

>>>>

% official NIPS keywords
\textbf{\\keywords}: mind-wandering, artificial intelligence, Markov decision processes

@@

\section{Introduction}
%
In the absence of external stimulation, the human brain is not at rest.
In the beginning of the 21st century,
brain imaging may have been the first technique
to allow for the discovery of a unique brain network that probably subserves
baseline mental activities
\citep{raichle2001pnas, randy2008, bzdok2015resting}.
The ``default mode network'' (DMN) continues
to metabolize large quantities of
oxygen and glucose energy to maintain
neuronal computation
during free-ranging thought
\citep{kenet2003spontaneously, fiser2004small}.
The baseline energy demand is only weakly modulated
at the onset of defined psychological tasks \citep{raichle_baseline}.
At its opposite, during sleep, the
decoupling of brain structures discarded the idea of the DMN being
only a passive network resonance and rather supported an
important role in sustaining conscious awareness \citep{horovitz2009decoupling}.

This \textit{dark matter of brain physiology}
\citep{raichle2006brain} begs the question
of the biological purpose underlying DMN activity.
What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
We propose that this set of some of the most advanced regions in
the association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control of human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex'' \citep{carhart2010default}.

@@

Since theories based on the FEP \citep{friston2010free,fristonAIorRL}
conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
given the environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a light-deprived closed space can be used
  as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would inevitably
identify the sensory-stimulus-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.

@@

\section{Conclusion}
Which brain function could be important enough
for the existence and survival of the human species
to justify constantly high energy costs?
MDPs motivate an attractive
formal account how the human association cortex
might implement multi-sensory representation and control of the environment to
optimize the organism's interaction with the world.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the view of a Markovian sequential decision process,
human behavior unfolds by integrating action outcomes from stored past events
and extrapolation to
upcoming events for guiding action choice in the present context.
MDPs also provide a mathematical formalism how
opportunity in the environment can be recursively exploited
when confronted with challenging decisions.
%
This functional interpretation may be more compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world representation
by injecting stochasticity into the recall of past
actions and the estimation of action outcomes may explain why
highly accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.

The proposed artificial intelligence perspective
on DMN biology is hence practically computable and
yields falsifiable neuroscientific hypotheses.
%
Neuroscience experiments could be designed that operationalize
the set of action, value, and state variables that govern
the behavior of intelligent RL agents.
%
At the least, we propose an alternative vocabulary to
describe and interpret experimental findings in neuroscience studies
on the DMN situated at top of the brain network hierarchy.
%
Ultimately,
DMN activity may instantiate a holistic integration
ranging from real experience over purposeful dreams to anticipated futures
for continued refinement of the organism's fate.

@@

% \appendix
% 
% \begin{remark}
%   Given a system $(\mathcal S, \mathcal A, r, p, \mu)$ satisfying all the axioms for an MDP
%   except the Markov property, we can always transform it into an MDP by considering a compounded
%   version of the states
%   $$S_t \leftarrow (S_0,A_0,R_0,S_1,A_1,S_2\ldots,S_t,A_{t-1},R_{t-1},S_{t})$$ made of the system's
%   history up to an including time $t$.
% \end{remark}
% 
% \section{More on RL in the brain}
% \subsection{Does hippocampal replay equal inverse reinforcement-learning?}
% %\begin{mdframed}
%   Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
%   an MDP (called a \textit{teacher's demonstration}), can we figure out what is the
%   (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
%   \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
%   (e.g., linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
%   For instance, given traces of motor actions from an adult to grab a cup on a table,
%   can an observing child figure out what ``langragian'' functional is being minimized by the former?
%   How? Can they reproduce this optimal behavior?
% %\end{mdframed}
% Such questions are of course pertinent to our decision-making theory for the DMN.
% In the general artificial case, the problem has been extensively studied and partially solved by
% \citep{abbeel2004}. They are also been rigorously studied in general optimal control literature under the name
% ``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...
% 
% IRL is is suited for problems in which it's hard to define what the reward function could be
% (e.g.,  car-driving, drone maneuvering, etc.) ...
% 
% \section{Free-energy principles!}
% The so-called free-energy principle in its present form (including notions like ``generative density'', ``recognition density'', etc.) can be traced back to works of Dayan \& Hinton \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial directional deep belief-net (DBN).
% In this subsection, we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in building their Helmholtz machine. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For instance, the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial) wake-sleep algorithm to train the underlying Helmholtz machine, etc.
% 
% % \begin{mdframed}
% %   Though our presentation of free-energy principles is thermodynamic in flavor, a completely equivalent
% %   information-theoretic formulation is of course possible, simply by exploiting the information-theoretic interpretation of entropy.
% % \end{mdframed}
% 
% \subsection{Helmholtz free-energy and the generative model}
% \begin{table}[H]
%   \begin{tabular}{p{2cm}|p{11cm}}
%          \hline
%          \textbf{symbol}    & \textbf{description}  \\ \hline
%          $\langle X\rangle_p$ & Expectation (a.k.a average, a.k.a mean value) of the
%          random quantity $X$ w.r.t to the probability density $p$, formally defined by $\langle E\rangle_p := \sum_{z}p(z)X(z)$.\\ \hline
%          $\mathcal H(p)$ & Information-theoretic entropy of a probability density $p$, formally defined by $\mathcal H(p) := -\sum_{z}p(z)\log(p(z))$,
%           with the usual convention $0 \log(0) := 0$.\\ \hline
%          $D_{KL}(q||p)$ & The Kullback-Leibler divergence between the probability densities $q$ and $p$ respectively, formally defined by $D_{KL}(q||p) := \sum_{z}q(z)\log(q(z)/p(z))$.\\ \hline
%              $\x$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's internal state $b$ and sensory inputs $s$, i.e., $\x = (s, b).$ \\ \hline
%              $\z$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
%              $p_G(.|\x)$ & Generative density for ...\\ \hline
%          $p_R(.|\x)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
%          $F_G(\x)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\x$. This measures the surprise incured upon observing $\x$ generated by the model $G$.\\ \hline
%          $F^R_G(\x)$ & Variational Helmholtz free-energy from model $G$
%           to $R$.  Note that $F^G_G = F_G$.\\ \hline
%   \end{tabular}
%   \caption{Table of notations.}
% \end{table}
% Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
% \begin{eqnarray}
%   \begin{split}
%     \text{generative surprise } &= -\log(p_G(\x)) = -\log(p_G(\x)) \times 1 = -\log(P_G(\x))\sum_{\z}p_G(\z |\x)\\
%     &= -\sum_{\z}p_G(\z, \x)\log(p_G(\x))
%     =-\sum_{\z}p_G(\z |\x)\log(p_G(\z, \x)/p_G(\z|\x))\\
%     &= \sum_{\z}p_G(\z |\x)\log(p_G(\z|\x))-\sum_{\z}p_G(\z |\x)\log  (p_G(\z, \x))\\
%     &= -\langle \log  (p_G(., \x)) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x))\\
%     &= \langle E_G(., \x) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x))
%   \end{split}
%   \label{eq:helm}
% \end{eqnarray}
% where $E_G(\z, \x)$ is the energy at \textit{macrostate} $\z$ of a fictive thermodynamic system defined by setting
% \begin{equation}
%   E_G(\z, \x) := -\log(p_G(\z, \x)),
%   \label{eq:gibbs}
% \end{equation}
% %% In the above, the variable $\z$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
% The last quantity in \eqref{eq:helm} is nothing but \textit{Helmholtz free-energy} (at unit temperature!), defined formally by
% \begin{equation}
%   F_G(\x) := \langle E_G(., \x) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x)).
% \end{equation}
% %% , modelled as the physical system for which $p_G(\z|\x)$ dictates the occupation probabilities of macrostates $\z$
% Thus,
% \begin{fact}
%   Generative surprise and generative Helmholtz free-energy are different views on exactly the same object.
% \end{fact}
% 
% The goal of the brain is then to optimize over the generative model $G$: to iteratively or analytically modify the generative density $p_G(.|\x)$, so as to minimize surprise. It turns out that a direct attempt to attack this optimization problem by gradient descent on the free-energy $F_G(\x)$ is furtile: the parameter update steps are not ``very clean'', and require rather cumbersome and heavy computations. A workaround is then to introduce a second density $p_R(.|\x)$ called a \textit{recognition} density to work in tandem with the generative density $p_G(.|\x)$, as a trick for doing approximate inference. The former dreams / fantacizes whilst the latter tries to generate sensations which match these dreams! This primal-dual idea, first proposed in Hinton et al. 1995, is at the heart of the general free-energy principle that we will introduce shortly.
% 
% \subsection{Variational Helmholtz free-energy and the bottom-up recognition sub-model}
% In this subsection, we will present an insightful upper bound for the generative surprise (i.e., generate Helmholtz free-energy), called the \textit{variational} (Helmholtz) free-energy. As an avant-gout of what is to come shorty, let's just note that the well-known \textit{free-energy principle} is simply a workaround whereby the minimization surprise (intractable) is replaced with the  minimization a carefully chosen upper bound thereof.
% 
% Invoking \eqref{eq:gibbs} and applying Bayes rule, we get the Gibbs
% distribution
% \begin{equation}
%   p_G(\z|\x) = \frac{p_G(\z|\x)}{p_G(\x)} = \frac{\exp(-E_G(\z, \x))}{Z_G(\x)} = \frac{\exp(-E_G(\z, \x))}{Z_G(\x)},
% \end{equation}
% where $Z_G(\x) := \log(p_G(\x)) = \sum_{\z'}\exp(-E_G(\z',\x))$, the normalizing \textit{partition function} for the model \ref{eq:gibbs}.
% Whence, for any macrostate $\z$, we have $p_G(\x) = Z_G(\x) = \exp(-E_G(\z, \x)) / p_G(\z|\x)$, and so it holds that
% \begin{equation}
% F_G(\x)\overset{\eqref{eq:helm}}{=} -\log(p_G(\x)) = -\log(Z_G(\x)) = E_G(\z, \x) + \log(p_G(\z|\x)).
% \end{equation}
% Now, in the above equation, the LHS only depends on the generative model $G$ and the data point $\x$:
% it doesn't depend on the hidden variable $\z$, etc. So, taking expectations w.r.t an arbitrary
% density\footnote{conditioning in $P_R(.|\x)$ is because this density is selected from a world in which the sensory inputs and internal brain
%   state vector $\x$ is assumed already observed.} $p_R(.|\x)$ yields
% \begin{equation}
%   \begin{split}
%     F_G(\x) &= -\log(Z_G(\x)) = \langle E_G(., \x)\rangle_{P_R(.|\x)} + \sum_{\z}p_R(\z|\x)\log(p_G(\z|\x))\\
%     &= \langle E_G(., \x)\rangle_{P_R(.|\x)} - \mathcal H(p_R(.|\x)) - \sum_{\z}p_R(\z|\x)\log(p_R(\z|\x)/p_G(\z|\x))\\
%     &= F^R_G(\x) - D_{KL}(P_R(.|\x) || P_G(.|\x)),
%   \end{split}
%   \label{eq:fe}
% \end{equation}
% where $F^R_G(\x)$ is the \textit{variational} Helmholtz free-energy from $R$ to $G$ defined by
% \begin{equation}
%   F^R_G(\x) := \langle E_G(., \x)\rangle_{P_R(.|\x)} - \mathcal H(p_R(.|\x))
% \end{equation}
% and $D_{KL}(P_R(.|\x) || P_G(.|\x))$ is the Kullback-Leibler divergence between the $p_R(.|\x)$ and the generative density $p_G(.|\x)$. Note that $F^G_G = F_G$.
% 
% \subsection{A general free-energy principle}
% We can resume the situation as follows\footnote{Where we have used the fact that KL divergence is always nonnegative.}:
% \begin{mdframed}
% \begin{equation}
%   \begin{split}
%     \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
%     &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{D_{KL}(P_R(.|\x) || P_G(.|\x))}_{\text{complexity}} \\
%     &\le F^R_G(\x),
%     \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z
%     \end{split}
% \end{equation}
% \end{mdframed}
% %% In fact, we have the following theorem
% %% \begin{theorem}
% %%   It holds that
% %%   $$
% %%   \text{Minimal generative surprise = minimal (Helmholtz) variational free-energy},$$
% %%   i.e
% %%   \begin{equation}
% %%     \min_{G}F_G(\x) = \min_{G,R}F^R_G(\x)
% %%     \end{equation}
% %% \end{theorem}
% 
% \subsection{Helmholtz machines and the wake-sleep algorithm}
% % \begin{mdframed}
%   \textbf{Assumption:} In both generative and recognition components of the network, there is conditional independence of  neurons in the same layer, given the data (i.e., input from lower more primitive layers). Precisely
% $$
% p_G(\z^{(l)}|\x) = \Pi_{k=1}^{h_l}p_G(\z_k^{(l)} | \x),\hspace{1cm}
% p_R(\z^{(l)}|\x) = \Pi_{k=1}^{h_l}p_R(\z_k^{(l)} |\x)
% $$
% % \end{mdframed}
% 
% \subsection{Friston's active-inference and agency}
% \label{sec:friston}
% This is nothing but an application of the Dayan's wake-sleep algorithm for training a Helmholtz machine model of the brain...
% 
% The following critics can be made:
% \begin{itemize}
%   \item
%     As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
%   \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
%   \item Also, note that the wake-sleep algorithm doesn't do backprop! This is due to
%     technical difficulty in getting derivatives of loss function w.r.t
%     recognition weights $\W^R$).
